Local Type Inference for Polarised System F with Existentials

### 2024-04-16 oopsla24/tex/01-introduction.tex
Edit intro and overview

@@

\paragraph{Type Inference Algorithm} We give a local type inference algorithm
    which supports both first-class existential and universal quantifiers, both
    of which can be instantiated impredicatively. To evade the undecidability
    results surrounding type inference for \systemf, we work in a variant of
    \CBPV \cite{levy2006:cbpv}, which lets us formulate a subtyping relation
    which is still decidable.

>>>>

In this paper, we extend the local type inference algorithm to a language with
both universal and existential quantifiers, which can both be instantiated
impredicatively. This extended system is referred to as \fexists. The
combination of features in \fexists  broke a number of the invariants which
traditional type inference algorithms depend on and required us to invent
new algorithms which combine both unification and (surprisingly)
anti-unification.

@@

\paragraph{Type Inference Algorithm} We give a local type inference algorithm
    which supports both first-class existential and universal quantifiers, both
    of which can be instantiated impredicatively. To evade the undecidability
    results surrounding type inference for \systemf, we introduce \fexists---a
    variant of \CBPV \cite{levy2006:cbpv}, which lets us formulate a subtyping
    relation which is still decidable.

### 2024-04-16 oopsla24/tex/02-overview.tex
Edit intro and overview

Positive expressions correspond to values in the \CBPV system
\cite{levy2006:cbpv}, and negative expressions correspond to computations.
The difference between the computations and values is intuitively described 
in the slogan of \CBPV: ``a value is, a computation does''. 
In particular, an argument of a function is always a value (\ie positive), 
and the resulting type, as well as the function itself, 
is a computation (\ie negative).

@@

In general, any term and type of \systemf can be embedded into \fexists in
several ways, because at each moment one can choose the positive or negative
polarity of the subterm. The consistent term-level and type-level embeddings
(that preserve typing) will be covered in detail in
\cref{sec:rel-to-systemf}. Now, we discuss several examples of standard
\systemf terms and types and their polarization in \fexists.

@@

Let us assume that $[[Φ]]$ is a context containing the terms from \cref{fig:polarization-examples}.
Then the following types are inferrable in \fexists.

\paragraph{Application of a Polymorphic Function}
  A polymorphic function can be applied to a value of a concrete type.
  In this case, the quantified variables are instantiated to make the application type-check.
  For instance, the polymorphic $[[id]]$ can be applied to any positive type, 
  and the resulting type will be inferred by the algorithm:
  $$[[· ; Φ ⊢ let x = id(fortyTwo); return x : ↑Int]]$$

\paragraph{Polymorphic Self-application}
  Unlike predicative systems, \fexists
  permits self-application of polymorphic functions.
  For instance, $[[id]]$ can be applied to itself
  resulting in a polymorphic type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$:
  $$[[· ; Φ ⊢ let x = id(id); let y = x(x); return y : ↑↓(∀α⁺.α⁺ → ↑α⁺)]]$$

\paragraph{Non-example: Polymorphic Arguments}
The focus of our system is the \emph{local} type inference \cite{pierce2000:local},
which has certain limitations.
In particular, if an argument of a function is polymorphic,
its polymorphic variables are \emph{not} instantiated by the inference algorithm. 
For example, one would expect the following to be inferrable:
\mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
However, to infer this, $[[id]]$ must be instantiated to \mbox{$[[Int → ↑Int]]$}
which requires the information to be propagated from the neighboring branch 
of the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.

@@

The types of \fexists are given in \cref{fig:declarative-types}.
They are stratified into two syntactic 
categories (polarities): positive and negative,  
similar to the values and computations in the \CBPV system \cite{levy2006:cbpv}.
\begin{itemize}
\item [$-$] $[[na]]$ is a negative type variable, which can be taken from a context or introduced by $[[∃]]$.
\item [$-$] a function $[[iP → iN]]$ takes a value as input and returns a computation; 
\item [$-$] a polymorphic abstraction $[[∀pas.iN]]$ quantifies a computation over
  a list of positive type variables 
  $[[pas]]$. The polarities are chosen to follow the definition of functions.
\item [$-$] a shift $[[↑iP]]$ allows a value to be used as a computation, 
  which at the term level corresponds to a pure computation $[[return v]]$.
\item [$+$] $[[pa]]$ is a positive type variable, taken from a context or introduced by $[[∀]]$.
\item [$+$] $[[∃nas.iP]]$, symmetrically to $[[∀]]$, 
  binds negative variables in a positive type $[[iP]]$. 
\item [$+$] a shift $[[↓iN]]$, symmetrically to the up-shift, 
  thunks a computation, which at the term level corresponds to $[[ {c} ]]$.
\end{itemize}

\begin{figure}[h]
  \begin{minipage}[t]{0.6\textwidth}
    \begin{supertabular}{ll p{0.8\textwidth}}
      \ottiNInline\\
    \end{supertabular}
  \end{minipage}
  \begin{minipage}[t]{0.35\textwidth}
    \begin{supertabular}{ll p{0.8\textwidth}}
      \ottiPInline\\
    \end{supertabular}
  \end{minipage}

@@

\paragraph{Definitional Equalities}
For simplicity, we assume that alpha-equivalent terms are equal.
This way, we assume that substitutions do not capture bound variables.
Besides, we equate
$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
and lift these equations transitively and congruently 
to the whole system.

@@

@@

\begin{itemize}
    \item [$+$] $[[x]]$ denotes a term variable.
      Following the \CBPV stratification, we only have \emph{positive} (value)
      term variables;
    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
        or suspended computation;
    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
        positive and negative terms;
    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
        are standard lambda abstractions. Notice that we require
        the type annotation for the argument of $[[λ]]$;
    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
    \item [$-$] Applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
        $[[let x = v ( args ) ; c]]$ operate similarly to 
        the bind of a monad: they take a suspended computation $[[v]]$,
        apply it to a list of arguments, bind the result 
        (which is expected to be pure) to a variable $[[x]]$,
        and continue with a computation $[[c]]$.
        If the resulting type of the application is unique, 
        one can omit the type annotation, as in the second form:
        it will be inferred by the algorithm;
    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
        is the standard unpack of an existential type:
        expecting $[[v]]$ to be an existential type,
        it binds the packed negative types to a list of 
        variables $[[nas]]$, binds the body of the existential
        to $[[x]]$, and continues with a computation $[[c]]$.
\end{itemize}

@@

\begin{figure}[h]
  \begin{supertabular}{ll p{0.8\textwidth}}
    \ottcInline\\
    \ottvInline\\
  \end{supertabular}
  \caption{Declarative Terms of \fexists}
  \label{fig:declarative-terms}
\end{figure}

>>>>

Positive expressions correspond to \emph{values} in the \CBPV system, and
negative expressions correspond to \emph{computations}. The difference between
the computations and values is intuitively described in the motto of \CBPV: ``a
value is, a computation does''. In particular, an argument of a function is
always a value (\ie positive), and the resulting type, as well as the function
itself, is a computation (\ie negative). Analogously, the polymorphic
$\forall$ quantifies a computation over positive types.

@@

Generally, any term and type of \systemf can be embedded into \fexists in
multiple ways. This is because at any given point, one can opt for either the
positive or negative polarity of the subterm. We will delve into the consistent
term-level and type-level embeddings (\ie preserving typing) in more detail in
\cref{sec:rel-to-systemf}. \Cref{fig:polarization-examples} showcases the
polarized versions of the standard \systemf terms. Next, we will discuss several
examples (using the terms from \cref{fig:polarization-examples}) to illustrate
what can and cannot be inferred in \fexists.

@@

\paragraph{Application of a Polymorphic Function}
  A polymorphic function can be applied to a value of a concrete type. In this
  case, the quantified variables are instantiated to make the application
  type-check. For instance, the polymorphic $[[len]]$ can be applied to a list
  of any positive type, and the resulting type will be inferred by the
  algorithm:
  $$[[· ; Φ ⊢ let x = len([one,two,three]); return x : ↑Int]]$$

\paragraph{Polymorphic Self-application}
  Unlike predicative systems, \fexists permits self-application of polymorphic
  functions. For instance, $[[id]]$ can be applied to itself resulting in a
  suspended polymorphic computation of type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$, which, in
  turn, can be applied to itself:
  $$[[· ; Φ ⊢ let x = id(id); let y = x(x); return y : ↑↓(∀α⁺.α⁺ → ↑α⁺)]]$$

\paragraph{Non-example: Polymorphic Arguments}
The focus of our system is the \emph{local} type inference
\cite{pierce2000:local}, which has certain limitations. In particular, the
instantiation does not happen when the polymorphic term is in the argument
position. For example, one would expect the following to be inferrable:
\mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
However, to infer this, $[[id]]$ must be instantiated to \mbox{$[[Int → ↑Int]]$}
which requires the information to be propagated from the neighboring branch of
the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.

@@

The types of \fexists are given in \cref{fig:declarative-types}.
They are stratified into two syntactic 
categories (polarities): positive and negative,  
similar to the values and computations in the \CBPV system.

\begin{figure}
  \begin{minipage}[t]{0.6\textwidth}
    \begin{supertabular}{ll p{0.8\textwidth}}
      \ottiNInline\\
    \end{supertabular}
  \end{minipage}
  \begin{minipage}[t]{0.35\textwidth}
    \begin{supertabular}{ll p{0.8\textwidth}}
      \ottiPInline\\
    \end{supertabular}
  \end{minipage}

@@

\begin{itemize}
\item [$\pm$] $[[na]]$ and $[[pa]]$
  denote negative and positive type variables, respectively.
  The variables are either introduced by the corresponding quantifiers ($[[∀]]$ and $[[∃]]$)
  or declared in the type context;
\item [$-$] a function $[[iP → iN]]$ takes a value as input and returns a computation; 
\item [$-$] similarly to functions, a polymorphic abstraction $[[∀pas.iN]]$ quantifies a computation over
  a list of positive type variables $[[pas]]$;
\item [$+$] dually, $[[∃nas.iP]]$, binds negative variables in a positive type $[[iP]]$;
\item [$-$] an upshift $[[↑iP]]$ allows a value to be used as a computation, 
  which at the term level corresponds to a pure computation $[[return v]]$;
\item [$+$] dually, a downshift $[[↓iN]]$ thunks a computation, which at the
  term level corresponds to $[[ {c} ]]$.
\end{itemize}

\paragraph{Definitional Equalities}
For simplicity, we assume that alpha-equivalent terms are equal.
This way, we assume that substitutions do not capture bound variables.
Besides, we assume the quantification is associative by
(congruently) equating $[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$.

@@

\begin{figure}
  \begin{supertabular}{ll p{0.8\textwidth}}
    \ottcInline\\
    \ottvInline\\
  \end{supertabular}
  \caption{Declarative Terms of \fexists}
  \label{fig:declarative-terms}
\end{figure}

@@

\begin{itemize}
    \item [$+$] $[[x]]$ denotes a term variable.
      Following the \CBPV stratification, we only have \emph{positive} (value)
      term variables;
    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
        or suspended computation;
    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
        positive and negative terms;
    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
        are standard lambda abstractions. Notice that we require
        the type annotation for the argument of $[[λ]]$;
    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
    \item [$-$] computational let form $[[ let x : iP = c; c']]$ 
      operates similarly to a monadic bind.
      It binds the result of a computation $[[c : ↑iP]]$ 
      to a variable $[[x : iP]]$, and continues with a computation $[[c']]$;
    \item [$-$] applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
        $[[let x = v ( args ) ; c]]$ can be viewed as a special case of the
        computational let, when the first computation is a function application
        $[[v]] ([[args]])$.
        They take a suspended computation $[[v]]$, apply it to a list of
        arguments, bind the result (which is expected to be pure) to a variable
        $[[x]]$, and continue with a computation $[[c]]$. If the resulting type
        of the application is unique, one can omit the type annotation, as in
        the second form: it will be inferred by the algorithm;
    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
        is the standard unpack of an existential type:
        expecting $[[v]]$ to be an existential type,
        it binds the packed negative types to a list of 
        variables $[[nas]]$, binds the body of the existential
        to $[[x]]$, and continues with a computation $[[c]]$.
\end{itemize}

@@
